Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-04-02 paper/paper.tex
Backpedaling on shift to superposition semantics. Causes too many problems. Still have results to prove.
Conflicts:

	paper/paper.tex

At each step during the initial presentation and development, we
evaluated the implementation on a set of benchmarks. The highlighted
benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.  For the
step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item it proves to be the \emph{least} improved benchmark of the
  complete suite considered in section~\ref{sec:eval}, and thus it
  serves as a good sanity check and lower-bound for each of the
  optimization techniques considered.
\end{enumerate}

@@

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{alloc}$, $\mathit{tick}$,
  $\mathit{push}$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
&&\mcntr' =\tick(\mcntr) \\
&&\maddr  =\alloc(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\push_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation from that state.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

{\bf FIXME} use $\traces$

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and it is deterministic even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

For clarity, we will exhibit non-widened semantics unless a technique
specifically requires it.

\subsection{Store-allocated results}
\label{sec:baselineeval}

The final approximation we make to get to our point of departure is
store-allocating results of application sub-expressions. 
The AAM approach stops at the previous optimization. 
However, the {\tt fn} continuation stores a value, and this makes the space of
continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
Having the space of continuations grow linearly with the size of the program 
will drop the overall complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate addresses to store the values of each
subexpression of an application. The continuations have to store these
addresses to remember where to get argument values from, though
sometimes your allocation strategy may allow you to elide that
representation.

The new evaluation rules follow (where $\mcntr' = \tick(\mlab,\mcntr)$).
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

@@

All techniques are precision preserving under reasonable
assumptions. We will discuss these assumptions as the caveats arise.

@@

The following example illustrates
the essence of abstract compilation effect:
\[
\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
makes the following transitions:
\begin{align}
& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
\machstep\; &
\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_{f1},\maddr_{a1},\maddr_1}}
\\
\machstep\; &
\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_{f2},\maddr_{a2},\maddr_2}}
\\
\machstep\; &
\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3}} % {\mexp_2}
\\
\machstep\; &
\spchoice
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\saddr{\menv(\mvar)}}}
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\superposition{\msto_4(\menv(\mvar))}}}
\end{align}
where \(
\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).

Whereas abstract compilation gives us in one step:
\begin{align*}
\compile{\mexp}^\mcntr(\menv,\msto_0,\mkont) = 
\spchoice
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\saddr{\menv(\mvar)}}}
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\superposition{\msto'_4(\menv(\mvar))}}}
\end{align*}
where \(
\msto'_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
[\maddr_2 \mapsto \kar{\compile{\mexp_3},\menv,\maddr_1}],
[\maddr_3 \mapsto \kar{\compile{\mexp_2},\menv,\maddr_2}]\).

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
{\tt ev} states. The next step is to change reduction rules that
create {\tt ev} states to instead call these functions. Figure
\ref{fig:caam} shows the modified reduction relation. The only change
from the previous semantics is that $\ev{}$ state construction is
replaced by calling the compiled expression.

\begin{figure}
\begin{align*}
\compile{\_} &: \Expr \to \Env \times \Store \times \Kont \to \State \\
\compile{\svar\mvar} &=
 \lambda(\menv,\msto,\mkont) .
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}}
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_f,\maddr_a,\maddr_\mkont})
\\
&\setlength\arraycolsep{5pt}\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) 
\\&
\maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
\end{array}
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \push_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}

>>>>

At each step during the initial presentation and development, we
evaluated the implementation on a set of benchmarks. The highlighted
benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.  For the
step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item its improvement is about median in the benchmark suite
  considered in section~\ref{sec:eval}, and thus it serves as a good
  sanity check for each of the optimization techniques considered.
\end{enumerate}

@@

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{alloc}$, $\mathit{tick}$,
  $\mathit{push}$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\push_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

Conceptually, the AAM approach calls for computing an analysis as a
graph exploration: (1) start with an initial state, and (2) compute
the transitive closure of the transition relation from that state. All
visited states are potentially reachable in the concrete, and all
paths through the graph are possible traces of execution.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph. We elide the construction of the set of edges in this paper.

@@

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and it is deterministic even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

For clarity, we will exhibit non-widened semantics unless a technique
specifically requires it. The final step to the baseline takes the
complexity from quartic to cubic in the monovariant case.

\subsection{Store-allocate all values}
\label{sec:baselineeval}

The final approximation we make to get to our point of departure is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the {\tt fn}
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to store this address
to remember where to get values from, though sometimes your allocation
strategy may allow you to elide that representation.

The new evaluation rules follow (where $\mcntr' = \tick(\mstate)$).
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

@@

All pragmatic improvments are precision preserving (form complete
abstractions), but the semantic changes are not in some cases, for reasons we
will describe. We did not observe the precision differences in our evaluation.

@@

The following example illustrates
the essence of abstract compilation effect:
\[
\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
makes the following transitions:
\begin{align}
& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
\machstep\; &
\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_{f1},\maddr_{a1},\maddr_1}}
\\
\machstep\; &
\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_{f2},\maddr_{a2},\maddr_2}}
\\
\machstep\; &
\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3}} % {\mexp_2}
\\
\machstep\; &
\spchoice
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
\end{align}
where \(
\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).

Whereas abstract compilation gives us in one step:
\begin{align*}
\compile{\mexp}^\mcntr(\menv,\msto_0,\mkont) = 
\spchoice
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto'_4}}
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
\end{align*}
where \(
\msto'_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
[\maddr_2 \mapsto \kar{\compile{\mexp_3},\menv,\maddr_1}],
[\maddr_3 \mapsto \kar{\compile{\mexp_2},\menv,\maddr_2}]\).

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
{\tt ev} states. The next step is to change reduction rules that
create {\tt ev} states to instead call these functions. Figure
\ref{fig:caam} shows the modified reduction relation. The only change
from the previous semantics is that $\ev{}$ state construction is
replaced by calling the compiled expression. For notational coherence,
we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
$\lambda(\mathit{args} \ldots, \mcntr)$ and
$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
\mcntr)$.

\begin{figure}
\begin{align*}
\compile{\_} &: (\Expr \to \Env \times \Store \times \Kont \times \Counter) \to \State \\
\mcntr' &= \tick(\mlab,\menv,\msto,\mcntr) \\
\compile{\svar\mvar} &=
 \lambda^\mcntr(\menv,\msto,\mkont) .
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}}
\\
\compile{\slit\mlit} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile{\mexp},\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
\\
&\setlength\arraycolsep{5pt}
\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont)\end{array}
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \push_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Abstract compilation}
\label{fig:compile}
\end{figure}
